The Galactic Delay-Time Distribution of Type Ia Supernovae:\\ A Chemical Evolution Perspective

### 2023-09-19 src/tex/ms.tex
[showyourwork] overleaf sync

Studies of galactic chemical evolution typically assume a single form for the SN Ia DTD. This is the first comprehensive look at the DTD in a model for GCE which includes the effects of radial migration.

@@

The origin of the [$\alpha$/Fe] bimodality is debated. \citet{Schonrich2009-RadialMixing} argue that it arises due to radial migration (churning) even with a smooth star formation rate. \citet{Johnson2021-Migration} use a smooth SFR and a prescription for radial migration but are unable to reproduce the bimodality. \citet{Spitoni2021-TwoInfall} (definitely earlier papers too) argue that two distinct infall episodes can explain the bimodality.

@@

@@

We limit our sample to red giant branch and red clump stars with high-quality spectra; Table \ref{tab:sample-selection} lists our selection criteria, which produce a final sample of \variable{output/sample_size.txt}stars.

@@

191,179 stars in our sample have astroNN ages and 71,684 have ages from \citet{Leung2023-Ages}.

@@

\begin{deluxetable*}{Ccl}
    \tablecaption{A summary of parameters and their fiducial values for our multi-zone chemical evolution models. We omit parameters which have been left unchanged from \citet{Johnson2021-Migration}; see their Table 1 for details.\label{tab:multizone-parameters}}
    \tablehead{
        \colhead{Quantity} & \colhead{Value} & \colhead{Description}
    }
    \startdata
        t_D                 & 40 Myr                                & Minimum SN Ia delay time \\
        \Delta R_{\rm gal}  & Equation \ref{eq:radial-migration}    & Change in radius due to stellar migration (see Appendix \ref{app:migration}) \\
        z                   & Equation \ref{eq:sech-pdf}            & Distance from Galactic midplane at present day (see Appendix \ref{app:migration}) \\
        f_{\rm EB}(t|R_{\rm gal})   & Equation \ref{eq:earlyburst-ifr}  & Time-dependence of the infall rate in the early-burst model \\
        f_{\rm TI}(t|R_{\rm gal})   & Equation \ref{eq:twoinfall-ifr}   & Time-dependence of the infall rate in the two-infall model \\
        \uptau_{*,\rm EB}(t|R_{\rm gal})  & Equation \ref{eq:earlyburst-conroy22} & Time-dependence of the SFE timescale in the early-burst model \\
    \enddata
\end{deluxetable*}

@@

@@

\begin{figure*}
    \centering
    \includegraphics{figures/ofe_bimodality_summary.pdf}
    \caption{The distributions of [O/Fe] along two slices of [Fe/H]: $-0.6\leq$[Fe/H]$<-0.4$ (blue dashed) and $-0.4\leq$[Fe/H]$<-0.2$ (red solid). \textit{Top row:} results from five multi-zone simulations which assume the late-bust SFH but different DTD models. \textit{Bottom row}: the first four panels compare the four SFH models, all assuming an exponential DTD with $\tau=1.5$ Gyr. The bottom-right panel plots data from APOGEE DR17 for reference. All panels contain stars within the Galactic region defined by $7\leq R_{\rm gal}<9$ kpc and $0\leq|z|<2$ kpc. The maximum of each distribution is normalized to 1.}
    \label{fig:ofe-bimodality}
    \script{ofe_bimodality_summary.py}
\end{figure*}

@@

\begin{figure}
    \centering
    \includegraphics{figures/age_ofe_sfh.pdf}
    \caption{A comparison of the age--[O/Fe] relation between multi-zone simulations with different SFH models. All assume an exponential DTD with $\tau=1.5$ Gyr. Each panel plots a random mass-weighted sample of 10,000 star particles in the Solar neighborhood ($7\leq R_{\rm gal}<9$ kpc, $0\leq|z|<0.5$ kpc) color-coded by the Galactocentric radius at birth. A Gaussian scatter has been applied to all points based on the median [O/Fe] error from APOGEE DR17 and the median age error from \citet{Leung2023-Ages} (see Table \ref{tab:uncertainties}). Black squares represent the mass-weighted median age of star particles within bins of [O/Fe] with a width of 0.05 dex, and the horizontal black error bars encompass the 16th and 84th percentiles. Red triangles represent the median age from \citet{Leung2023-Ages} in bins of [O/Fe], and the horizontal red error bars encompass the 16th and 84th percentiles. For clarity, bins which contain less than 1\% of the total mass (in the simulations) or total number of stars (in the data) are not plotted.}
    \label{fig:age-ofe-sfh}
    \script{age_ofe_sfh.py}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics{figures/age_ofe_dtd.pdf}
    \caption{A comparison of the age--[O/Fe] relation between multi-zone simulations with different DTD models. All assume an inside-out SFH. Each row contains star particles from a different bin in $|z|$, with stars closest to the midplane in the bottom row and stars farthest from the midplane in the top row. In all panels stars are limited to the Solar annulus ($7\leq R_{\rm gal}<9$ kpc), and the layout of each panel is as in Figure \ref{fig:age-ofe-sfh}.}
    \label{fig:age-ofe-dtd}
    \script{age_ofe_dtd.py}
\end{figure*}

@@

@@

When a star particle is created by VICE at initial radius $R_{\rm form}$, we sample its total radial migration distance $\Delta R$ from a Gaussian with a width described by Equation \ref{eq:radial-migration}, and we sample its final midplane distance $z_{\rm final}$ from the distribution described by Equation \ref{eq:sech-pdf} with a width given by Equation \ref{eq:scale-height}. The star particle migrates to its final radius $R_{\rm final}$ in a similar manner to the ``diffusion'' case from \citet{Johnson2021-Migration}, but with a time dependence $\propto t^{1/3}$. This produces distributions of $R_{\rm final}$ and $z_{\rm final}$ which are similar to the analogue migration case for all but the oldest stars.

Figure \ref{fig:radial-migration} compares the distributions of $R_{\rm final}$ in bins of $R_{\rm form}$ and age between the analogue and Gaussian migration schema. The

>>>>

Studies of galactic chemical evolution typically assume a single form for the SN Ia DTD. This is the first comprehensive look at the DTD in a multi-zone GCE model which includes the effects of radial migration.

@@

Studies which have noted the [$\alpha$/Fe] bimodality: \citet{Furhmann1998-NearbyStars,Hayden2015-ChemicalCartography}. The origin of the [$\alpha$/Fe] bimodality is debated. \citet{Schonrich2009-RadialMixing} argue that it arises due to radial migration (churning) even with a smooth star formation rate. \citet{Johnson2021-Migration} use a smooth SFR and a prescription for radial migration but are unable to reproduce the bimodality. \citet{Spitoni2021-TwoInfall} (definitely earlier papers too) argue that two distinct infall episodes can explain the bimodality.

@@

\begin{deluxetable*}{Cccl}
    \tablecaption{A summary of parameters and their fiducial values for our chemical evolution models. We omit parameters which have been left unchanged from \citet{Johnson2021-Migration}; see their Table 1 for details.\label{tab:multizone-parameters}}
    \tablehead{
        \colhead{Quantity} & \colhead{Fiducial Value(s)} & \colhead{Section} & \colhead{Description}
    }
    \startdata
        % Multi-zone simulation parameters
        R_{\rm gal}     & [0, 20] kpc   & & Galactocentric radius \\
        \delta R_{\rm gal}  & 100 pc    & & Width of each concentric ring \\
        \Delta R_{\rm gal}  & N/A       & \ref{app:migration} & Change in orbital radius due to stellar migration \\
        p(\Delta R_{\rm gal}|\tau,R_{\rm form}) & Equation \ref{eq:radial-migration}    & \ref{app:migration} & Probability density function of radial migration \\
        z                   & [-3, 3] kpc                & \ref{app:migration} & Distance from Galactic midplane at present day \\
        p(z|\tau,R_{\rm final}) & Equation \ref{eq:sech-pdf}            & \ref{app:migration} & Probability density function of Galactic midplane distance\\
        \Delta t        & 10 Myr    & & Time-step size \\
        t_{\rm max}     & 13.2 Gyr  & & Maximum simulation time \\
        n               & 8         & & Number of stellar populations formed per ring per time-step \\
        R_{\rm Ia}(t)   & Equations \ref{eq:powerlaw-dtd}--\ref{eq:triple-dtd}  & \ref{sec:dtd-models}  & Delay time distribution of Type Ia supernovae \\
        t_D             & 40 Myr    & \ref{sec:dtd-models}  & Minimum SN Ia delay time \\
        M_{\rm gas,0}   & 0         & & Initial gas mass \\
        % Nucleosynthetic yields
        y_{\rm O}^{\rm CC}  & 0.015     & & CCSN yield of O    \\
        y_{\rm Fe}^{\rm CC} & 0.0012    & & CCSN yield of Fe   \\
        y_{\rm O}^{\rm Ia}  & 0         & & SN Ia yield of O       \\
        y_{\rm Fe}^{\rm Ia} & 0.00214   & & SN Ia yield of Fe \\
        f_{\rm IO}(t|R_{\rm gal})   & Equation \ref{eq:insideout-sfh}   & & Time-dependence of the star formation rate in the inside-out model \\
        f_{\rm LB}(t|R_{\rm gal})   & Equation \ref{eq:lateburst-sfh}   & & Time-dependence of the star formation rate in the late-burst model \\
        f_{\rm EB}(t|R_{\rm gal})   & Equation \ref{eq:earlyburst-ifr}  & & Time-dependence of the infall rate in the early-burst model \\
        \uptau_{*,\rm EB}(t|R_{\rm gal})  & Equation \ref{eq:earlyburst-conroy22} & & Time-dependence of the SFE timescale in the early-burst model \\
        f_{\rm TI}(t|R_{\rm gal})   & Equation \ref{eq:twoinfall-ifr}   & & Time-dependence of the infall rate in the two-infall model \\
    \enddata
\end{deluxetable*}

@@

We limit our sample to red giant branch and red clump stars with high-quality spectra. Table \ref{tab:sample-selection} lists our selection criteria, which largely follow from \citet{Hayden2015-ChemicalCartography}. This produces a final sample of \variable{output/sample_size.txt}stars.

@@

\citet{Leung2023-Ages} use a variational encoder-decoder network to estimate the age of APOGEE giants without contamination from age-abundance correlations. Their sample does not experience the $\sim10$ Gyr plateau that affects ages from astroNN \citep{Mackereth2019-astroNN-Ages}. We use an age uncertainty cut of 40\% per the recommendations of \citet{Leung2023-Ages}, which produces a total sample of \variable{output/age_sample_size.txt}APOGEE stars with age estimates.

% 191,179 stars in our sample have astroNN ages and 71,684 have ages from \citet{Leung2023-Ages}.

@@

@@

Away from the midplane, there is a clear trend of a higher average stellar [O/Fe] at small $R_{\rm gal}$ and lower [O/Fe] at large $R_{\rm gal}$. However, the APOGEE data show a clear dichotomy between inner-galaxy, high-$\alpha$ stars and outer-galaxy, low-$\alpha$ stars, with few in between. In the inside-out and late-burst SFHs, the shift from high-$\alpha$ to low-$\alpha$ is much more gradual and the trough at [O/Fe]$\approx0.2$ does not appear. The two-infall SFH appears to produce \textit{three} modes, at [O/Fe]$\approx 0.0$, $0.2$, and $0.4$, with the first and last showing the most variation with $R_{\rm gal}$ at high latitudes. The early-burst SFH produces the closest match to the data by far: the low-$\alpha$ sequence away from the midplane shows increasing contribution with larger radii, with the innermost radii containing very few low-$\alpha$ stars. However, the high-$\alpha$ sequence contains substantial contribution from stars at large radii, which is not seen in the data.

@@

A striking feature of the [O/Fe] distribution in the Milky Way is the distinct separation into two components, the high- and low-$\alpha$ sequences.

\begin{figure*}
    \centering
    \includegraphics{figures/ofe_bimodality_summary.pdf}
    \caption{The distributions of [O/Fe] along two slices of [Fe/H]: $-0.6\leq$[Fe/H]$<-0.4$ (blue dashed) and $-0.4\leq$[Fe/H]$<-0.2$ (red solid). \textit{Top row:} results from five multi-zone simulations which assume the late-bust SFH but different DTD models. \textit{Bottom row}: the first four panels compare the four SFH models, all assuming an exponential DTD with $\tau=1.5$ Gyr. The bottom-right panel plots data from APOGEE DR17 for reference. All panels contain stars within the Galactic region defined by $7\leq R_{\rm gal}<9$ kpc and $0\leq|z|<2$ kpc. The distributions of stars in $|z|$ The maximum of each distribution is normalized to 1 and the vertical scale is consistent across all panels.}
    \label{fig:ofe-bimodality}
    \script{ofe_bimodality_summary.py}
\end{figure*}

@@

\begin{figure}
    \centering
    \includegraphics{figures/age_ofe_sfh_alt.pdf}
    \caption{A comparison of the age--[O/Fe] relation between multi-zone simulations with different SFH models. All assume an exponential DTD with $\tau=1.5$ Gyr. Each panel plots a random mass-weighted sample of 10,000 star particles in the Solar neighborhood ($7\leq R_{\rm gal}<9$ kpc, $0\leq|z|<0.5$ kpc) color-coded by the Galactocentric radius at birth. A Gaussian scatter has been applied to all points based on the median [O/Fe] error from APOGEE DR17 and the median age error from \citet{Leung2023-Ages} (see Table \ref{tab:uncertainties}). Black squares represent the mass-weighted median age of star particles within bins of [O/Fe] with a width of 0.05 dex, and the horizontal black error bars encompass the 16th and 84th percentiles. Red triangles represent the median age from \citet{Leung2023-Ages} in bins of [O/Fe], and the horizontal red error bars encompass the 16th and 84th percentiles. For clarity, bins which contain less than 1\% of the total mass (in the simulations) or total number of stars (in the data) are not plotted.}
    \label{fig:age-ofe-sfh}
    \script{age_ofe_sfh_alt.py}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics{figures/age_ofe_dtd_alt.pdf}
    \caption{A comparison of the age--[O/Fe] relation between multi-zone simulations with different DTD models. All assume an inside-out SFH. Each row contains star particles from a different bin in $|z|$, with stars closest to the midplane in the bottom row and stars farthest from the midplane in the top row. In all panels stars are limited to the Solar annulus ($7\leq R_{\rm gal}<9$ kpc), and the layout of each panel is as in Figure \ref{fig:age-ofe-sfh}.}
    \label{fig:age-ofe-dtd}
    \script{age_ofe_dtd_alt.py}
\end{figure*}

@@

\citet{Palicio2023-AnalyticDTD} compare seven different DTDs, including the analytical SD, DD-WIDE, and DD-CLOSE, as well as the empirical two-population and $t^{-1}$ power-law DTDs, in a one-zone model with a two-infall SFH. [Note that they use some different parameters for the analytical models than we do). Critically for our purposes, and in contrast to previous studies of the two-infall model \citep[e.g.,][others]{Chiappini1997-TwoInfall,Spitoni2021-TwoInfall}, the authors include the effects of outflows, although they explore mass-loading values of $\eta=0.2-0.8$ which is significantly lower than in our simulations.

@@

When a star particle is created by VICE at initial radius $R_{\rm form}$, we sample its total radial migration distance $\Delta R$ from a Gaussian with a width described by Equation \ref{eq:radial-migration}, and we sample its final midplane distance $z_{\rm final}$ from the distribution described by Equation \ref{eq:sech-pdf} with a width given by Equation \ref{eq:scale-height}. The star particle migrates to its final radius $R_{\rm final}$ in a similar manner to the ``diffusion'' case from \citet{Johnson2021-Migration}, but with a time dependence $\propto \Delta t^{1/3}$. This produces distributions of $R_{\rm final}$ and $z_{\rm final}$ which are similar to the analogue migration case for all but the oldest stars.

\citet{Okalidis2022-AurigaMigration} studied the stellar migration in the Auriga simulations and found that the migration strength had both a dependence on age and formation radius. They also found that in strongly-barred galaxies, the migration is stronger overall but has a slower timestep evolution than diffusion ($\Delta t^{0.5}$).

Figure \ref{fig:radial-migration} compares the distributions of $R_{\rm final}$ in bins of $R_{\rm form}$ and age between the analogue and Gaussian migration schema.
